### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix alignment of \subst
Desc@ICFP: fix alignment of \replace
Desc@ICFP: ispell

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specializing} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{(\PLAM{x}{\EnumT{E}}\Desc)}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together
the declaration of \(\Desc\) and its internal copy, we have made
it appear in its own type. Hardwaired as a \emph{fait accompli},
this creates no regress, although one must assume the definition
to recheck it.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          (\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})})\: 
          \SHED
}\]
%
We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

>>>>

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{(\PLAM{x}{\EnumT{E}}\Desc)}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together
the declaration of \(\Desc\) and its internal copy, we have made
it appear in its own type. Hardwired as a \emph{fait accompli},
this creates no regress, although one must assume the definition
to recheck it.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          (\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})})\: 
          \SHED
}\]
%
We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To \\
         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

### 2010-04-02 papers/icfp-2010-desc/paper_type_theory.tex
added split, tweaked the examples
Desc@ICFP: restore type-checking as a floating figure
Desc@ICFP: type-synthesis floating
Desc@ICFP: ispell

We adapt this idea, yielding a \emph{type
propagation} system, whose purpose is to elaborate compact
\emph{expressions} into the terms of our underlying type theory, much
as in the definition of Epigram
1~\cite{mcbride.mckinna:view-from-the-left}.  We divide expressions
into two syntactic categories: $\exprIn$ into which types are pushed,
and $\exprEx$ from which types are extracted. In the
bidirectional spirit, the $\exprIn$ are subject to type
\emph{checking}, while the $\exprEx$---variables and elimination
forms---admit type \emph{synthesis}. We embed $\exprEx$ into
$\exprIn$, demanding that the synthesized type coincides with the type
proposed. The other direction---only necessary to apply
abstractions or project from pairs---takes a type annotation.

@@

Type synthesis is the \emph{source} of types. It follows the \(\exprEx\)
syntax, delivering both the elaborated term and its type. Terms
and expressions never mix: e.g., for application, we
instantiate the range with the \emph{term} delivered by checking the
argument \emph{expression}. Hardwired operators are checked as variables.
\input{figure_type_synthesis}

%\begin{figure}
%\input{figure_type_synthesis}
%\caption{Type synthesis}
%\label{fig:type-synthesis}
%\end{figure}

@@

Dually, type checking judgments
%(Fig.~\ref{fig:type-checking})
are \emph{sinks} for types. From an
$\exprIn$ and a type pushed into it, they elaborate a low-level
term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
%
\[\stkc{
\Gamma\Vdash\propag{e}{\pull{t}{T}} \Rightarrow
  \Gamma\vdash t:T \\
\Gamma\Vdash\push{\propag{e}{t}}{T} \Rightarrow
  \Gamma\vdash t:T
}\]

@@

%\begin{figure}
\input{figure_type_checking}
%\caption{Type checking}
%\label{fig:type-checking}
%\end{figure}

>>>>

We adapt this idea, yielding a \emph{type
propagation} system, whose purpose is to elaborate compact
\emph{expressions} into the terms of our underlying type theory, much
as in the definition of Epigram
1~\cite{mcbride.mckinna:view-from-the-left}.  We divide expressions
into two syntactic categories: $\exprIn$ into which types are pushed,
and $\exprEx$ from which types are extracted. In the
bidirectional spirit, the $\exprIn$ are subject to type
\emph{checking}, while the $\exprEx$---variables and elimination
forms---admit type \emph{synthesis}. We embed $\exprEx$ into
$\exprIn$, demanding that the synthesised type coincides with the type
proposed. The other direction---only necessary to apply
abstractions or project from pairs---takes a type annotation.

@@

Type synthesis (Fig.~\ref{fig:type-synthesis}) is the \emph{source} of
types. It follows the \(\exprEx\) syntax, delivering both the
elaborated term and its type. Terms and expressions never mix: e.g.,
for application, we instantiate the range with the \emph{term}
delivered by checking the argument \emph{expression}. Hardwired
operators are checked as variables.

\begin{figure}
\input{figure_type_synthesis}
\caption{Type synthesis}
\label{fig:type-synthesis}
\end{figure}

@@

Dually, type checking judgments (Fig.~\ref{fig:type-checking})
are \emph{sinks} for types. From an
$\exprIn$ and a type pushed into it, they elaborate a low-level
term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
%
\[\stkc{
\Gamma\Vdash\propag{e}{\pull{t}{T}} \Rightarrow
  \Gamma\vdash t:T \\
\Gamma\Vdash\push{\propag{e}{t}}{T} \Rightarrow
  \Gamma\vdash t:T
}\]

@@

\begin{figure}
\input{figure_type_checking}
\caption{Type checking}
\label{fig:type-checking}
\end{figure}
